Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2022-06-03 analysis/paper/shceof.Rmd
Renderiza

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 46,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

>>>>

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 42,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

### 2022-06-03 analysis/paper/shceof.tex
Renderiza

\begin{abstract}
We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
The leading cEOF represents variabilty of a zonal wave 1 troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes.
One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere.
Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2.
\\
\keywords{
        Southern Hemisphere circulation \and
        Teleconnections \and
        Pacific South American Mode \and
        Southern Annular Mode \and
        Stratosphere \and
    }

@@

Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the Real cEOF1 is associated with a centre located over the Ross Sea.
The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.

At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase (e.g. Fogt and Marshall 2020).

Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.

@@

Therefore, by virtue of being the most common phase, the Imaginary cEOF2 explains more variance than the Real cEOF2.
Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the Imaginary cEOF2 always leading the one representing the Real cEOF2.

This phase preferences is in agreement with Irving and Simmonds (2016), who found a bimodal distribution to PSA-like variability (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).

@@

In the case of Imaginary cEOF2, the correlation between SAM and Imaginary cEOF2 for the troposphere is associated to the asymmetric variability of the SAM.
Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022).

@@

The correlation between the absolute magnitude of the ONI index and the cEOF2 amplitude is 0.45 (CI: 0.17 -- 0.66).
However, this relationship is mostly driven by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which coincide with the three years with strongest cEOF2 magnitude (not shown).
If those years are removed, the correlation becomes non-significant (0.04 (CI: -0.28 -- 0.35)).
Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain might be evident only in particularly intense ENSO events, for which there are few data.

@@

The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the Southern Hemisphere is also explored.
Figure~\ref{fig:pp-t2m-r2} shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real and Imaginary cEOF2 components (column 2).
The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.~ \ref{fig:pp-t2m-r2}a.1).

@@

Our method is similar to Goyal et al. (2022) as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2.
Both indices are likely very similar (a detailed comparison is out of scope for this paper) but we believe that cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.

@@

\leavevmode\vadjust pre{\hypertarget{ref-goyal2022}{}}%
Goyal, Rishav, Martin Jucker, Alex Sen Gupta, and Matthew H. England. 2022. {``A New Zonal Wave 3 Index for the {Southern Hemisphere}.''} \emph{Journal of Climate} -1 (May): 1--25. \url{https://doi.org/10.1175/JCLI-D-21-0927.1}.

>>>>

\begin{abstract}
The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
In this study we analyse the joint variability of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
The leading cEOF represents variability of a zonal wave 1 in the stratosphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes.
One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere.
Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2.
As a result, the methodology proposed in this study provides a deeper understanding of the zonally asymmetric springtime extratropical SH circulation.
\\
\keywords{
        Southern Hemisphere circulation \and
        Teleconnections \and
        Pacific South American Mode \and
        Southern Annular Mode \and
        Stratosphere \and
    }

@@

Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the Real cEOF1 is associated with a centre located over the Ross Sea.
The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.
At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase (e.g. Fogt and Marshall 2020).
Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.

@@

Therefore, by virtue of being the most common phase, the Imaginary cEOF2 explains more variance than the Real cEOF2.
Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the Imaginary cEOF2 always leading the one representing the Real cEOF2.
This phase preferences is in agreement with Irving and Simmonds (2016), who found a bimodal distribution to PSA-like variability (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).

@@

In the case of Imaginary cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM.
Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022).

@@

The correlation between the absolute magnitude of the ONI index and the cEOF2 amplitude is 0.45 (CI: 0.17 -- 0.66).
However, this relationship is mostly driven by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which coincide with the three years with strongest cEOF2 magnitude (not shown).
If those years are removed, the correlation becomes non-significant (0.04 (CI: -0.28 -- 0.35)).
Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain and might be only evident in very strong ENSO events, that are scarce in the historical observational record.

@@

The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the SH was also explored.
Figure~\ref{fig:pp-t2m-r2} shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real and Imaginary cEOF2 components (column 2).
The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.~ \ref{fig:pp-t2m-r2}a.1).

@@

Our method is similar to Goyal et al. (2022) as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2.
Both indices are likely very similar (a detailed comparison is out of scope for this paper) but cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.

@@

\leavevmode\vadjust pre{\hypertarget{ref-goyal2022}{}}%
Goyal, Rishav, Martin Jucker, Alex Sen Gupta, and Matthew H. England. 2022. {``A New Zonal Wave 3 Index for the {Southern Hemisphere}.''} \emph{Journal of Climate} -1 (May): 1--25. \url{https://journals.ametsoc.org/view/journals/clim/aop/JCLI-D-21-0927.1/JCLI-D-21-0927.1.xml}.
